# Cluster Analysis on Travel Discrimination
\
Dataset explanation:\
**Variables:**\
**Continuos**\
- Q1_ = Travel frequency\
- Q6_15 : checkin experience rate\
- Q6_18 : fly experience rate\
**Categorical**\
- Q15 = Gender\
- Q17 = Race\
- Q18 = Religion\

\
## Library
```{r,warning=FALSE,message=FALSE}
library(readr)
library(readxl)
library(tidyverse)
library(corrplot)
library(ggfortify)
library(FactoMineR)
library(factoextra)
library(gplots)
library(ggpubr)
library(magrittr)
```

## Read the dataset

```{r}
travel <- read_excel("data_ordination.xlsx")
head(travel)
```

Dataset contains 231 rows and 14 columns which is still messy. Thus, we'll conduct some data preprocessing steps.\

## DATA PREPROCESSING

```{r}
# First, drop two first rows. 
travel <- travel %>% 
  slice(-c(1,2))

# Select used columns
travel_df <- travel[c(1,5,6,9,12,13,14)]
```
\
```{r}
# CHECK MISSING VALUE----
# Count the missing values by column wise
print("Count of missing values by column wise")
sapply(travel_df, function(x) sum(is.na(x)))

# Missing value imputation
# Since our data contains 46 missing value, let's impute with mode
# Function to see mode
calc_mode <- function(x){
  
  # List the distinct / unique values
  distinct_values <- unique(na.omit(x))
  
  # Count the occurrence of each distinct value
  distinct_tabulate <- tabulate(match(x, distinct_values))
  
  # Return the value with the highest occurrence
  distinct_values[which.max(distinct_tabulate)]
}

# Impute missing value----
travel_df <- travel_df %>% 
  mutate(across(everything(), ~replace_na(.x, calc_mode(.x))))
```
\
```{r}
# Rename column name
travel_df_clean <- travel_df %>% 
       rename(respondent_id =1, travel_frequency = 2, checkin_exp = 3,
              fly_exp = 4, gender = 5, race=6,
              religion = 7)
head(travel_df_clean)
```
\

```{r}
# CONVERT DATA TYPE----
# Convert all variables into integer
# Convert column 2 to 6 to numeric
travel_df_clean[,2:4] <- lapply(travel_df_clean[,2:4], as.numeric)
travel_df_clean[,5:7] <- lapply(travel_df_clean[,5:7], as.factor)
travel_df_clean[,5:7] <- lapply(travel_df_clean[,5:7], as.integer)
head(travel_df_clean)
```
## Correspondence Analysis
### Data Exploration
```{r}
# set respondent_id as index
travel_df_clean <- travel_df_clean %>% column_to_rownames(., var = "respondent_id")
head(travel_df_clean)
```

```{r}
# convert data into contingency table
df1 = travel_df_clean[,-c(2)]
dt = as.table(as.matrix(df1))

# graph
balloonplot(t(dt[1:15,]), main = "Travel Discriminations", xlab ="",
            ylab = "", label= FALSE, show.margins = FALSE)
```

```{r}
res_ca <- CA(df1, graph = T)
print(res_ca)
```

```{r}
# eigen values
eig_val <- get_eigenvalue(res_ca)
eig_val
```

```{r, warning=F, message=F}
# Statistical Significance 1
chisq <- chisq.test(travel_df_clean)
chisq
```

```{r}
# statistical significance 2
summary(res_ca)
```

### Scree Plot
```{r}
fviz_screeplot(res_ca, addlabels = TRUE, ylim = c(0, 50))
```
```{r}
fviz_screeplot(res_ca) +
 geom_hline(yintercept=25, linetype=2, color="red")
```
### BiPlot
```{r}
# repel= TRUE to avoid text overlapping
#options(ggrepel.max.overlaps = Inf)
fviz_ca_biplot(res_ca, repel = TRUE)
```

### Graph of row Variables
```{r}
row <- get_ca_row(res_ca)
row
```

#### Access the Row Component
```{r}
# Coordinates
head(row$coord)
# Cos2: quality on the factore map
head(row$cos2)
# Contributions to the principal components
head(row$contrib)
```

#### Coordinates Row Points
```{r}
fviz_ca_row(res_ca, repel = TRUE)
```
#### Quality of Representation of Rows
```{r}
# Color by cos2 values: quality on the factor map
fviz_ca_row(res_ca, col.row = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE)
```

```{r}
# Change the transparency by cos2 values
fviz_ca_row(res_ca, alpha.row="cos2")
```

#### Corrplot
```{r}
corrplot(row$cos2[1:15,], is.corr=FALSE)
```

```{r}
# Cos2 of rows on Dim.1 and Dim.2
fviz_cos2(res_ca, choice = "row", axes = 1:2)
```
#### Contributions of rows to the dimensions
```{r}
corrplot(row$contrib[1:15,], is.corr=FALSE) 
```

```{r}
# Contributions of rows to dimension 1
fviz_contrib(res_ca, choice = "row", axes = 1, top = 10)
# Contributions of rows to dimension 2
fviz_contrib(res_ca, choice = "row", axes = 2, top = 10)
```

```{r}
# Total contribution to dimension 1 and 2
fviz_contrib(res_ca, choice = "row", axes = 1:2, top = 10)
```

```{r}
fviz_ca_row(res_ca, col.row = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
             repel = TRUE)
```

```{r}
# Change the transparency by contrib values
fviz_ca_row(res_ca, alpha.row="contrib",
             repel = TRUE)
```
### Graph of column Variables
```{r, warning=FALSE, message=FALSE}
col <- get_ca_col(res_ca)
col
```

```{r}
# Coordinates of column points
head(col$coord)
# Quality of representation
head(col$cos2)
# Contributions
head(col$contrib)
```

```{r}
fviz_ca_col(res_ca)
```

```{r}
fviz_ca_col(res_ca, col.col = "cos2", 
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE)
```

```{r}
fviz_cos2(res_ca, choice = "col", axes = 1:2)
```

```{r}
fviz_contrib(res_ca, choice = "col", axes = 1:2)
```
#### Symmetric Biplot
```{r}
fviz_ca_biplot(res_ca, repel = TRUE)
```
### Asymmetric Biplot
```{r}
fviz_ca_biplot(res_ca, 
               map ="rowprincipal", arrow = c(TRUE, TRUE),
               repel = TRUE)
```
### Contribution Biplot
```{r}
fviz_ca_biplot(res_ca, map ="colgreen", arrow = c(TRUE, FALSE),
               repel = TRUE)
```
### Dimension Description
```{r}
# Dimension description
res_desc <- dimdesc(res_ca, axes = c(1,2))
# Description of dimension 1 by row points
head(res_desc[[1]]$row)
# Description of dimension 1 by column points
head(res_desc[[1]]$col)
```

```{r}
# Description of dimension 2 by row points
head(res_desc[[2]]$row)
# Description of dimension 1 by column points
head(res_desc[[2]]$col)
```

## Multidimensional Scaling (MDS)
```{r}
# Cmpute MDS
mds <- df1 %>%
  dist() %>%          
  cmdscale() %>%
  as_tibble()
colnames(mds) <- c("Dim.1", "Dim.2")

# Plot MDS
ggscatter(mds, x = "Dim.1", y = "Dim.2", 
          label = rownames(df1),
          size = 1,
          repel = TRUE)
```
#### Create 3 groups k-means clustering
```{r}
# K-means clustering
clust <- kmeans(mds, 2)$cluster %>%
  as.factor()
mds <- mds %>%
  mutate(groups = clust)
# Plot and color by groups
ggscatter(mds, x = "Dim.1", y = "Dim.2", 
          label = rownames(df1),
          color = "groups",
          palette = "jco",
          size = 1, 
          ellipse = TRUE,
          ellipse.type = "convex",
          repel = TRUE)
```
## Cannonical Correspondence Analysis
```{r, warning=FALSE, message=FALSE}
library(vegan)
```

```{r}
t1 = as.matrix(df1)
example_NMDS=metaMDS(t1,k=2,trymax=100)
```
```{r}
stressplot(example_NMDS)
```

```{r}
plot(example_NMDS)
```

```{r}
# First, create a vector of color values corresponding of the 
# same length as the vector of treatment values
treat=c(rep("Treatment1",5),rep("Treatment2",5))
colors=c(rep("red",5),rep("blue",5))
ordiplot(example_NMDS,type="n")
#Plot convex hulls with colors baesd on treatment
for(i in unique(treat)) {
  ordihull(example_NMDS$point[grep(i,treat),],draw="polygon",
   groups=treat[treat==i],col=colors[grep(i,treat)],label=F) } 
orditorp(example_NMDS,display="species",col="red",air=0.01)
orditorp(example_NMDS,display="sites",col=c(rep("green",5),
  rep("blue",5)),air=0.01,cex=1.25)
```
