---
title: "Factor Analysis Analysis"
author: "Nadia Ahmad"
date: "2023-04-22"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
---
## Library
```{r,warning=FALSE,message=FALSE}
library(readr)
library(readxl)
library(tidyverse)
library(corrplot)
library(ggfortify)
library(psych)
library(gplots)
library(ggpubr)
library(magrittr)
library(REdaS)
library(nFactors)
```

## Read the dataset

```{r}
travel <- read_excel("factor_analysis_data.xlsx")
head(travel)
```

Dataset contains 228 rows and 13 columns which is still messy. Thus, we'll conduct some data preprocessing steps.\

## DATA PREPROCESSING
```{r}
# CHECK MISSING VALUE----
# Count the missing values by column wise
print("Count of missing values by column wise")
sapply(travel, function(x) sum(is.na(x)))

# Missing value imputation
# Since our data contains 46 missing value, let's impute with mode
# Function to see mode
calc_mode <- function(x){
  
  # List the distinct / unique values
  distinct_values <- unique(na.omit(x))
  
  # Count the occurrence of each distinct value
  distinct_tabulate <- tabulate(match(x, distinct_values))
  
  # Return the value with the highest occurrence
  distinct_values[which.max(distinct_tabulate)]
}

# Impute missing value----
travel_df <- travel %>% 
  mutate(across(everything(), ~replace_na(.x, calc_mode(.x))))
```
\

```{r}
# CONVERT DATA TYPE----
# Convert all variables into integer
# Convert column 2 to 6 to numeric
travel_df[,c(2, 9:13)] <- lapply(travel_df[,c(2, 9:13)], as.integer)
head(travel_df)
```
## Statistics Summary
```{r}
summary(travel_df)
```
## Correlation Matrix
```{r}
cor(travel_df)
```

## Kaiser-Meyer-Oklin Test (KMO)
```{r}
KMO(travel_df)
```
KMO overall should be .60 or higher to proceed with factor analysis. Thus, we will drop variables that have MSA less than 0.6
```{r}
# create new clean df
travel_df_clean <- travel_df[,-c(2:3, 9:12)]
head(travel_df_clean)
```

```{r}
# correlation matrix
cor(travel_df_clean)
```

```{r}
KMO(travel_df_clean)
```

```{r}
# Barlett Test
bart_spher(travel_df_clean)
```

## Principal Component Analysis
```{r}
pca1 = princomp(travel_df_clean, scores=TRUE, cor=TRUE)
summary(pca1)
```

```{r}
loadings(pca1)
```
### Scree Plot
```{r}
plot(pca1)
```

```{r}
screeplot(pca1, type="line", main="Scree Plot") 
```

## Determine Number of Factors to Retain
```{r}
ev <- eigen(cor(travel_df_clean)) 
ap <- parallel(subject=nrow(travel_df_clean), var=ncol(travel_df_clean), rep=100,cent=.05)
nS <- nScree(x=ev$values, aparallel=ap$eigen$qevpea)
plotnScree(nS)
```

## Exploratory Factorial Analysis
```{r}
fa.parallel(travel_df_clean)
```
## Principal Component
```{r}
fa1 <- principal(travel_df_clean, nfactors=2, rotate="none")
fa1
```
## Principal Axis Factor Analysis
```{r}
fa2 <- factor.pa(travel_df_clean, nfactors=2, rotate="none")
fa2
```
## Principal Component Factor Analysis with Rotation
```{r}
fa.none <- factanal(travel_df_clean, factors = 2, rotation = "none")
fa.varimax <- factanal(travel_df_clean, factors = 2, rotation = "varimax")
fa.promax <- factanal(travel_df_clean, factors = 2, rotation = "promax")

par(mfrow = c(1,3))
plot(fa.none$loadings[,1], 
     fa.none$loadings[,2],
     xlab = "Factor 1", 
     ylab = "Factor 2", 
     ylim = c(-1,1),
     xlim = c(-1,1),
     main = "No rotation")
abline(h = 0, v = 0)

plot(fa.varimax$loadings[,1], 
     fa.varimax$loadings[,2],
     xlab = "Factor 1", 
     ylab = "Factor 2", 
     ylim = c(-1,1),
     xlim = c(-1,1),
     main = "Varimax rotation")

text(fa.varimax$loadings[,1]-0.08, 
     fa.varimax$loadings[,2]+0.08,
      colnames(travel_df_clean),
      col="blue")
abline(h = 0, v = 0)

plot(fa.promax$loadings[,1], 
     fa.promax$loadings[,2],
     xlab = "Factor 1", 
     ylab = "Factor 2",
     ylim = c(-1,1),
     xlim = c(-1,1),
     main = "Promax rotation")
abline(h = 0, v = 0)
```
```{r}
fa.varimax
```

## Exploratory Factor Analysis Using Minimum Residual
```{r}
faPC_min_res <- fa(r = travel_df_clean, nfactors = 2,rotate = "none")
print(faPC_min_res)
```

```{r}
faPC_fa <- fa(r = travel_df_clean, nfactors = 2, fm = "pa",rotate = "none")
print(faPC_fa)
```

```{r}
faPC_ml <- fa(r = travel_df_clean, nfactors = 2, fm="ml", rotate = "none")
print(faPC_ml)
```
## Model Comparison
```{r}
faPC_min_res$rms
faPC_fa$rms
faPC_ml$rms
```

```{r}
fa.diagram(faPC_min_res)
fa.diagram(faPC_ml)
fa.diagram(faPC_fa)
```

```{r}
faPC_ml_r <- fa(r = travel_df_clean, nfactors = 2, fm="ml", rotate = "varimax")
print(faPC_ml_r)
```
```{r}
fa.diagram(faPC_ml_r)
```

```{r}
root.cor.pa <- principal(travel_df_clean, nfactors = 2, rotate = 'varimax')
root.cor.pa
```

```{r}
biplot(root.cor.pa)
```