---
title: "MANOVA"
author: "Nadia Ahmad"
date: "2023-03-17"
output:
  word_document: default
  html_document:
    df_print: paged
---

# MANOVA Model on Travel Discrimination
\
Dataset explanation:\
**Dependent Variable:**\
- Q6_15 : checkin experience rate\
- Q6_18 : fly experience rate\
**Independent Variables:**\
**Categorical**\
- Q15 = Gender\
- Q17 = Race\
- Q18 = Religion\
**Continuos**\
- Q1_ = Travel frequency\
\
## Library
```{r,warning=FALSE,message=FALSE}
library(readr)
library(readxl)
library(tidyverse)
library(corrplot)
library(MASS)
library(mvtnorm)
library(MVN)
library(psych)
library(ggfortify)
library(ggpubr)
library(mvoutlier)
library(heplots)
library(biotools)
library(emmeans)
library(jmv)
library(vegan)
```

## Read the dataset

```{r}
travel <- read_excel("data_manova.xlsx")
head(travel)
```

Dataset contains 231 rows and 14 columns which is still messy. Thus, we'll conduct some data preprocessing steps.\

## DATA PREPROCESSING

```{r}
# First, drop two first rows. Next, filter only data that has 100 in progress
travel <- travel %>% 
  slice(-c(1,2))

# Select used columns
travel_df <- travel[c(5,6,9,12,13,14)]
```
\
```{r}
# CHECK MISSING VALUE----
# Count the missing values by column wise
print("Count of missing values by column wise")
sapply(travel_df, function(x) sum(is.na(x)))

# Missing value imputation
# Since our data contains 46 missing value, let's impute with mode
# Function to see mode
calc_mode <- function(x){
  
  # List the distinct / unique values
  distinct_values <- unique(na.omit(x))
  
  # Count the occurrence of each distinct value
  distinct_tabulate <- tabulate(match(x, distinct_values))
  
  # Return the value with the highest occurrence
  distinct_values[which.max(distinct_tabulate)]
}

# Impute missing value----
travel_df <- travel_df %>% 
  mutate(across(everything(), ~replace_na(.x, calc_mode(.x))))
```
\
```{r}
# Rename column name
travel_df_clean <- travel_df %>% 
       rename(travel_frequency = 1, checkin_exp = 2,
              fly_exp = 3, gender = 4, race=5,
              religion = 6)
head(travel_df_clean)
```
\

```{r}
# CONVERT DATA TYPE----
# Convert all variables into integer
# Convert column 2 to 6 to numeric
travel_df_clean[,1:3] <- lapply(travel_df_clean[,1:3], as.numeric)
travel_df_clean[,4:6] <- lapply(travel_df_clean[,4:6], as.factor)
head(travel_df_clean)
```

## 1. INTERACTION PLOT
#### Checkin Experience Rate
##### Gender vs Race
```{r}
interaction.plot(
  x.factor = travel_df_clean$race,
  trace.factor = travel_df_clean$gender,
  response = travel_df_clean$checkin_exp,
  type="b", col=c(1:4),
  ylab = "Checkin Experience Rate",
  xlab = "Races",
  trace.label = "Gender",
  main="Interaction Plot",
  leg.bty="o", leg.bg="beige", lwd=2,
)
```

##### Gender vs Religion
```{r}
interaction.plot(
  x.factor = travel_df_clean$religion,
  trace.factor = travel_df_clean$gender,
  response = travel_df_clean$checkin_exp,
  type="b", col=c(1:4),
  ylab = "Checkin Experience Rate",
  xlab = "Religion",
  trace.label = "Gender",
  main="Interaction Plot",
  leg.bty="o", leg.bg="beige", lwd=2,
)
```

##### Race vs Religion
```{r}
interaction.plot(
  x.factor = travel_df_clean$religion,
  trace.factor = travel_df_clean$race,
  response = travel_df_clean$checkin_exp,
  type="b", col=c(1:8),
  ylab = "Checkin Experience Rate",
  xlab = "Religion",
  trace.label = "Race",
  main="Interaction Plot",
  leg.bty="o", leg.bg="beige", lwd=2
)
```

#### fly Claim Experience Rate
##### Gender vs Race
```{r}
interaction.plot(
  x.factor = travel_df_clean$race,
  trace.factor = travel_df_clean$gender,
  response = travel_df_clean$fly_exp,
  type="b", col=c(1:4),
  ylab = "Flight Experience Rate",
  xlab = "Races",
  trace.label = "Gender",
  main="Interaction Plot",
  leg.bty="o", leg.bg="beige", lwd=2,
)
```

##### Gender vs Religion
```{r}
interaction.plot(
  x.factor = travel_df_clean$religion,
  trace.factor = travel_df_clean$gender,
  response = travel_df_clean$fly_exp,
  type="b", col=c(1:4),
  ylab = "Flight Experience Rate",
  xlab = "Religion",
  trace.label = "Gender",
  main="Interaction Plot",
  leg.bty="o", leg.bg="beige", lwd=2,
)
```

##### Race vs Religion
```{r}
interaction.plot(
  x.factor = travel_df_clean$religion,
  trace.factor = travel_df_clean$race,
  response = travel_df_clean$fly_exp,
  type="b", col=c(1:8),
  ylab = "Flight Experience Rate",
  xlab = "Religion",
  trace.label = "Race",
  main="Interaction Plot",
  leg.bty="o", leg.bg="beige", lwd=2
)
```

## 2. MANOVA
### ANOVA WITH INTERACTION : Gender and Religion as Factors
#### Dependent : checkin experience
```{r}
aov_1 <- aov(checkin_exp ~ gender * religion, data = travel_df_clean)
summary(aov_1)
```

#### Dependent : fly claim experience
```{r}
aov_2 <- aov(fly_exp ~ gender * religion, data = travel_df_clean)
summary(aov_2)
```

### ANOVA WITH INTERACTION : Gender and Race as Factors
#### Dependent : checkin experience
```{r}
aov_1_ <- aov(checkin_exp ~ gender * race, data = travel_df_clean)
summary(aov_1_)
```

#### Dependent : fly claim experience
```{r}
aov_2_ <- aov(fly_exp ~ gender * race, data = travel_df_clean)
summary(aov_2_)
```

### ANOVA WITH INTERACTION : Religion and Race as Factors
#### Dependent : checkin experience
```{r}
aov_1_1 <- aov(checkin_exp ~ religion * race, data = travel_df_clean)
summary(aov_1_1)
```

#### Dependent : fly claim experience
```{r}
aov_1_2 <- aov(fly_exp ~ religion * race, data = travel_df_clean)
summary(aov_1_2)
```

### TWO WAY MANOVA WITH INTERACTION : Gender and Race as Factors
```{r}
manova1 <- manova(cbind(checkin_exp, fly_exp) ~ gender*race, data=travel_df_clean)
summary(manova1)
summary.aov(manova1)
```

### TWO WAY MANOVA WITH INTERACTION : Gender and Religion as Factors
```{r}
manova2 <- manova(cbind(checkin_exp, fly_exp) ~ gender*religion, data=travel_df_clean)
summary(manova2)
summary.aov(manova2)
```

### TWO WAY MANOVA WITH INTERACTION : Gender and Religion as Factors
```{r}
manova3 <- manova(cbind(checkin_exp, fly_exp) ~ race*religion, data=travel_df_clean)
summary(manova3)
summary.aov(manova3)
```

### ASSUMPTION TEST : PILLAI
```{r}
summary(manova3,
        test = "Pillai")
```

### ASSUMPTION TEST : WILKS
```{r}
summary(manova3,
        test = "Wilks")
```
\
### ASSUMPTION TEST : HOTELLING-LAWLEY
```{r}
summary(manova3,
        test = "Hotelling-Lawley")
```

### ASSUMPTION TEST : ROY
```{r}
summary(manova3,
        test = "Roy")
```
Assume now, we'll go with religion and race as factors based on MANOVA result.\

## 3. CONTRAST 
### Univariate Contrast
#### Race
```{r}
M <- lm(cbind(checkin_exp, fly_exp) ~ race + religion, data = travel_df_clean)
anova(M)
mvpaircomp(M, factor1 = "race")
```
#### Religion
```{r}
M <- lm(cbind(checkin_exp, fly_exp) ~ race + religion, data = travel_df_clean)
anova(M)
mvpaircomp(M, factor1 = "religion")
```
### Multivariate Contrast
```{r}
MOats.lm <- lm(cbind(checkin_exp, fly_exp) ~ race + religion, data = travel_df_clean)
MOats.emm <- emmeans(MOats.lm, ~ race| rep.meas)
mvcontrast(MOats.emm, "consec", show.ests = TRUE) 
```
\
```{r}
MOats.lm2 <- lm(cbind(checkin_exp, fly_exp) ~ race + religion, data = travel_df_clean)
MOats.emm2 <- emmeans(MOats.lm, ~ religion| rep.meas)
mvcontrast(MOats.emm2, "consec", show.ests = TRUE) 
```

## 4. MANCOVA
#### Plot : travel_frequency vs checkin_exp
```{r}
plot(travel_df_clean$travel_frequency, travel_df_clean$checkin_exp, main="Scatterplot of Travel Frequency vs Check In Experience Rate",
   xlab="travel_frequency ", ylab="Check_in_exp ", pch=19)
```
#### Plot : travel_frequency vs flight_exp
```{r}
plot(travel_df_clean$travel_frequency, travel_df_clean$fly_exp, main="Scatterplot of Travel Frequency vs Flight Experience Rate",
   xlab="travel_frequency ", ylab="flight_exp ", pch=19)
```
Now we add travel_frequency as covariates.
\
```{r, warning=FALSE}
results <- mancova(data = travel_df_clean,
        deps = vars(checkin_exp, fly_exp),
        factors = vars(race, religion),
        covs = travel_frequency,
        multivar = list("pillai", "wilks", "hotel", "roy"), boxM = T,
        shapiro = T, qqPlot = T)
```

### MULTIVARIATE MANCOVA
```{r}
results$multivar
```

#### UNIVARIATE MANCOVA
```{r}
results$univar
```

## 5. MODEL ASSUMPTION
### A. BOX M
```{r}
results$assump$boxM
```

### B. SHAPIRO WILK TEST
```{r}
results$assump$shapiro
```

### C. QQ-PLOT
```{r}
results$assump$qqPlot
```

## 6. MRPP : Multiresponse permutation procedures
```{r}
y <- as.matrix(travel_df_clean[,2:3])
y_mat <- sqrt(y) # square root transformation
```
\
```{r, warning=F}
y_dist <- vegdist(y_mat)
```

```{r}
set.seed(123)
mrpp <- adonis2(y_dist~ travel_df_clean$race * travel_df_clean$religion, permutations = 1000)
mrpp
```
\
